---
title: "Exercise 1B: SpatialFeatureExperiment"
format: html
editor: source
editor_options:
  chunk_output_type: console
execute:
  warning: false
---

## Learning Objectives

By the end of this exercise, you will be able to:

-   Understand the structure of a `SpatialFeatureExperiment` object.
-   Explain when `SpatialFeatureExperiment` is useful compared with `SpatialExperiment`.
-   Access cell-level metadata and spatial coordinates from Xenium data.
-   Inspect geometry information such as cell or nucleus boundaries.
-   Compare the Xenium representation with the Visium HD representation from Exercise 1 A.

## Libraries

```{r day1-1b-SFE-1}
#| message: false
#| warning: false

library(SpatialFeatureExperiment)
library(SpatialExperiment)
library(Voyager)
library(ggplot2)
library(patchwork)
library(scuttle)
library(sfarrow) #JUNA: needs to be installed 
library(HDF5Array)
```

## Data for the course

We will work on a **Xenium** dataset from a human colorectal cancer study by [Oliveira _et al_](https://www.nature.com/articles/s41588-025-02193-3). The dataset contains normal adjacent tissue (NAT) and colorectal carcinoma (CRC) from 5 patients.

We will focus on a single Xenium slide from a patient called "P2", available from the [10X website](https://www.10xgenomics.com/platforms/visium/product-family/dataset-human-crc). This slide is a serial section of the sample used in `Exercise 1 A`, so the two datasets are biologically related but not the exact same physical tissue section. 

The full Xenium output is very large and includes cell boundaries, nucleus boundaries, transcript locations, morphology images, and expression matrices. For this exercise, for practical reasons we use a region of interest, selected to **approximately match the region of interest used in `Exercise 1 A`**, and only include some of the layers in the imported compact course folder. This makes the two exercises comparable while avoiding a full image-registration workflow during the practical. This region was chosen from the overview plot because it contains gland-like epithelial structures together with stromal or lower-density areas.


```{r day1-1b-SFE-2}
roi_xenium <- c(
  xmin = 2200,
  xmax = 4500,
  ymin = 2500,
  ymax = 4900
)
roi_xenium
```

We import the Xenium output with `flip = "none"` so the exercise uses the native Xenium coordinate orientation. The same visual region may appear with negative y coordinates if the geometry has been flipped during import with morphology images. Because the compact course folder does not include morphology images, we use the positive y coordinates shown above.

```{r day1-1b-SFE-3}
options(timeout = 600)

if (!dir.exists("data/Human_Colon_Cancer_P2/")) {
  dir.create("data", showWarnings = FALSE)
  download.file(
    url = "https://intro-spatial-transcriptomics-training.s3.eu-central-1.amazonaws.com/Human_Colon_Cancer_P2.tar.gz",
    destfile = "data/Human_Colon_Cancer_P2.tar.gz",
    mode = "wb"
  )

  untar(
    tarfile = "data/Human_Colon_Cancer_P2.tar.gz",
    exdir = "data/"
  )

  file.remove("data/Human_Colon_Cancer_P2.tar.gz")
} else {
  message("Data exists, please proceed to next steps!")
}
```
<!-- JUNA: I don't think we need again the code to download the data, is in exercise 1a (with correct link) -->


We import the compact Xenium output into a `SpatialFeatureExperiment` object:

```{r day1-1b-SFE-4}
sfe_full <- readXenium("data/Human_Colon_Cancer_P2/xenium/outs", flip = "none")

rownames(sfe_full) <- uniquifyFeatureNames(
  ID = rowData(sfe_full)$ID,
  names = rowData(sfe_full)$Symbol
)

dim(sfe_full)
range(spatialCoords(sfe_full)[, 1])
range(spatialCoords(sfe_full)[, 2])
```

Then we subset it to the region of interest:

```{r day1-1b-SFE-5}
sfe <- sfe_full[, spatialCoords(sfe_full)[, 1] >= roi_xenium[["xmin"]] &
  spatialCoords(sfe_full)[, 1] <= roi_xenium[["xmax"]] &
  spatialCoords(sfe_full)[, 2] >= roi_xenium[["ymin"]] &
  spatialCoords(sfe_full)[, 2] <= roi_xenium[["ymax"]]]

sfe
```

::: callout-important
## Exercise 1

Why is an object of the `SpatialFeatureExperiment` class used to store Xenium data?

Which information could not be represented by a simple spot- or bin-level `SpatialExperiment`?
:::

::: {.callout-tip collapse="true"}
## Answer

Xenium is an image-based spatial transcriptomics technology. In addition to a count matrix and cell coordinates, it can provide cell boundaries, nucleus boundaries, transcript locations, and morphology images.

`SpatialFeatureExperiment` extends `SpatialExperiment` with geometry support, making it better suited to storing and manipulating these spatial shapes.
:::

## Exploring the object

`SpatialFeatureExperiment` extends `SpatialExperiment`, which extends `SingleCellExperiment`, so familiar accessors still apply:

```{r day1-1b-SFE-6}
dim(sfe)
colData(sfe) |> head()
rowData(sfe) |> head()
spatialCoords(sfe) |> head()
```

::: callout-important
## Exercise 2

Check the dimensions and metadata of the `sfe` object.

Questions:

-   What do the columns represent?
-   What do the rows represent?
-   What differs compared to the Visium HD `spe` object in `Exercise 1 A`?
-   What is stored in the main assay?
:::

::: {.callout-tip collapse="true"}
## Answer

In this object, columns represent segmented cells, while rows represent genes. This differs from the Visium HD slide stored in a `SpatialExperiment` object, where columns represent spatial bins rather than segmented cells.

```{r day1-1b-SFE-7}
ncol(sfe)
nrow(sfe)
```

We notice that the number of features is small (541) as the Xenium technology measures a targeted panel rather than the whole transcriptome.

```{r day1-1b-SFE-7b}
assay(sfe)
```

The count matrix indicates the number of observed molecules for each of the 541 genes in each cell. This matrix is analogous to a UMI counts matrix in the Visium HD dataset.

:::

## Inspecting geometries

In addition to the slots from the `SpatialExperiment` object, one of the key features of `SpatialFeatureExperiment` is that it can store geometries associated with cells or features.

```{r day1-1b-SFE-8}
colGeometries(sfe)
```

::: callout-important
## Exercise 3

Inspect the available column geometries in the Xenium object. To which biological structures do they correspond to? In which class are stored these objects?
:::

::: {.callout-tip collapse="true"}
## Answer

The available geometries include cell centroids, cell boundaries, and nucleus boundaries. These are are commonly available for imaging-based spatial transcriptomics technologies, but the exact names depend on the reader functions and packages used to import the data.

```{r day1-1b-SFE-9}
colGeometries(sfe)[["centroids"]]
```
The geometries are stored using "Simple Features" objects, which allow to encode spatial data and can be easily manipulated with the [`sf` package](https://r-spatial.github.io/sf/index.html)

:::

## Visualizing Xenium cells

We can start with a simple scatterplot of the cells in the selected region of interest:

```{r day1-1b-SFE-10}
xy <- as.data.frame(spatialCoords(sfe))
names(xy)[1:2] <- c("x", "y")

ggplot(xy, aes(x = x, y = y)) +
  geom_point(size = 0.08, alpha = 0.4) +
  coord_fixed() +
  theme_bw() +
  labs(x = "x", y = "y")
```

::: callout-important
## Exercise 4

Compare this point-based visualization with the Visium HD plot from `Exercise 1 A`.

What does one point represent in each dataset?
:::

::: {.callout-tip collapse="true"}
## Answer

In the Visium HD exercise, each point or square represents a 16 um spatial bin. In the Xenium exercise, each point represents a segmented cell centroid. This is one of the key conceptual differences between binned sequencing-based spatial data and image-based cell-resolved spatial data.
:::

We can also visualize expression for a marker gene. For example, `PIGR` is present in the Xenium panel and shows a clear spatial pattern in this region. We renamed the rows to gene symbols after import, which is why we can refer to this marker as `"PIGR"` rather than by its Ensembl ID. At this stage we plot the number of molecules since log-normalized counts will be introduced later in the course. The `plotSpatialFeature()` function from the `Voyager` package is handy:

```{r day1-1b-SFE-11}
plotSpatialFeature(sfe, "PIGR", exprs_values = "counts")
```

::: callout-important
## Exercise 5

Choose another marker gene and visualize its spatial pattern.

How does the interpretation differ from plotting a gene on Visium HD bins?

:::
<!-- JUNA: If we want them to compare to Visium HD we should not delete environment in previous exercise, or load the spe object here as part of the solution -->



::: {.callout-tip collapse="true"}
## Answer

The Xenium signal is measured over segmented cells, while the Visium HD binned signal aggregates molecules within fixed spatial bins. Xenium can therefore represent cell-level heterogeneity more directly, but it measures a targeted panel rather than the whole transcriptome.
:::

## Exercise 6

Use the `plotSpatialFeature()` function to visualize the cell segmentation mask, colored by cell area, and the nuclei segmentation mask, colored by nucleus area

::: {.callout-tip collapse="true"}
## Answer

```{r day1-1b-SFE-12}
plotSpatialFeature(sfe, 
                   colGeometryName="cellSeg", 
                   features="cell_area") 

plotSpatialFeature(sfe, 
                   colGeometryName="nucSeg", 
                   features="nucleus_area") 
```
:::
1
## Save the object

```{r day1-1b-SFE-13}
dir.create("results/day1", showWarnings = FALSE, recursive = TRUE)

saveHDF5SummarizedExperiment(sfe,
  dir = "results/day1", prefix = "01.1b_sfe_", replace = TRUE,
  chunkdim = NULL, level = NULL, as.sparse = NA,
  verbose = NA
)
```

Clear your environment:

```{r day1-1b-SFE-14}
#| eval: false
rm(list = ls())
gc()
.rs.restartR()
```

::: callout-important
**Key Takeaways:**

-   `SpatialFeatureExperiment` is useful for geometry-rich spatial data.
-   Xenium observations are cell-resolved, while the Visium HD object in Exercise 1 A is bin-resolved.
-   The P2 CRC subsets are designed to be comparable in course scope and approximate tissue region, not identical at cell level.
:::
