---
title: "Exercise 2"
format: html
editor: source
editor_options: 
  chunk_output_type: console
execute:
  warning: false
---


## Quality control at cell-level

In this second exercise, we will focus on the critical step of quality control (QC) for sequencing-based spatial transcriptomics data. We will focus on the segmented Visium HD dataset used in `Exercise 1C`, so each observation is an individual cell with the selected region of interest. QC helps flag low-quality cells or technical artifacts before downstream analysis.

*NOTE:* A similar QC can be done per bin or aggregated bin if a binned analysis of Visium HD is performed.

## Learning objectives

By the end of this exercise, you will be able to:

- Calculate per-cell QC metrics.
- Identify local outliers based on various QC metrics.
- Detect spatial artifacts using `SpotSweeper`.
- Visualize QC metrics and detected artifacts.


## Libraries
```{r}
#| message: false
#| warning: false
#| output: false

library(SpatialExperiment)
library(SpatialFeatureExperiment)
library(Voyager)
library(HDF5Array)

library(SpotSweeper)
library(scuttle)
library(scrapper)
library(scater)
library(escheR) 

library(ggplot2)
library(ggside)
library(ggspavis)
library(patchwork)
```

## Calculate QC metrics

We will start by calculating QC metrics that are also commonly used in scRNA-seq data analysis: total UMI counts, number of detected genes, and percentage of UMIs originating from genes of the mitochondrial genome. Here these metrics are calculated per segmented cell and used to identify low-quality cells.

We will start by loading our `SpatialFeatureExperiment` object, which was saved in the previous exercise, and prepare it for quality control analysis.  


```{r}
# Load the SpatialExperiment object if not in the R session already
sfe <- loadHDF5SummarizedExperiment(dir="results/day1/", prefix="01.1_sfe_visium_")

# Identify mitochondrial genes (genes with symbol starting with "MT-")
is.mito <- rownames(sfe)[grepl("^MT-", rownames(sfe))]
```

<!--- NOTE: HDF5-backed saving/loading via saveHDF5SummarizedExperiment() is reliable for assays, but it can leave sf geometries with invalid external pointers after reload (GEOS/s2), which then crash functions like localOutliers() with “external pointer is not valid”. 
Need a solution or don't delete previous object 
--->

Now, we will calculate per-cell QC metrics using the `scrapper` package. The function `quickRnaQc.se` adds QC metrics, including mitochondrial gene percentage, to the `colData` of the `SpatialFeatureExperiment` object. Run the following code to compute these metrics and inspect the results:

```{r}
sfe <- quickRnaQc.se(sfe, subsets = list(Mito = is.mito))
```

::: callout-important
## Exercise 1

Check which metadata has been added to `colData`. Do you recognize the different metrics?
:::

:::{.callout-tip collapse="true"}
### Answer

```{r}
# Display the colData to see the newly added QC metrics.
colData(sfe)
```

- The `sum` column contains the total number of unique molecular identifiers (UMIs) for each cell
- The `detected` column contains the number of unique genes detected per cell
- The `subset.proportion.Mito` column contains the percentage of transcripts mapping to mitochondrial genes per cell
:::

### Global outlier detection

In order to detect low-quality cells and filter them out, QC methods adapted from scRNA-seq analysis can be applied. A simple option is to apply a fixed upper or lower threshold to a metric across all cells and remove the cells that do not pass the filtering criteria.

To set up these thresholds we can look for information from 10x Genomics and the community, for example https://github.com/10XGenomics/HumanColonCancer_VisiumHD/issues/28. Based on this discussion the authors of the OSTA book wrote (see "Remove bins overlaying empty tissue" section in https://lmweber.org/OSTA/pages/seq-workflow-visium-hd-bin.html#dependencies):

> Based on a discussion with the original author at 10x Genomics, in this dataset, 8 µm bins with a library size below 100 are removed. Since we expect a 4-fold increase in library size across all bins at 16 µm, we can set the filtering threshold to 400.

We can also have a look and plot distributions of the values to decide which is the best cutoff:

```{r}
# Plot library size vs. detected genes
p1 <- ggcells(sfe, aes(x=detected, y=sum)) + 
  geom_point() +
  geom_density_2d() +
  scale_x_log10() +
  scale_y_log10() +
  geom_xsidehistogram() + 
  geom_ysidehistogram() + 
  geom_vline(xintercept = 300, col="red", lty=2) +
  geom_hline(yintercept = 400, col="red", lty=2) +
  labs(y="Number of UMIs", x="Number of detected genes") +
  ggtitle("Library size vs. detected")

# Plot mito proportion vs. detected genes
p2 <- ggcells(sfe, aes(x=detected, y=subsets_Mito_percent)) + 
  geom_point() +
  geom_density_2d() +
  scale_x_log10() +
  geom_xsidehistogram() + 
  geom_ysidehistogram() + 
  geom_vline(xintercept = 300, col="red", lty=2) +
  geom_hline(yintercept = 20, col="red", lty=2) +
  labs(y="% mitochondrial reads", x="Number of detected genes") +
  ggtitle("Mito proportion vs. genes detected")

p1 + p2
```

And further check how many cells would be removed with the given thresholds:

```{r}
# Flag cells based on fixed thresholds
sfe$qc_lib_size <- sfe$sum < 400
sfe$qc_detected <- sfe$detected < 300
sfe$qc_mito_prop <- sfe$subsets_Mito_percent > 20

# Tabulate number of cells flagged by each
qc <- grep("^qc", names(colData(sfe)))
sapply(colData(sfe)[qc], table)
```

::: callout-important
## Exercise 2
What do you think about this initial filtering?

Have a look at the mitochondrial percentage values on the tissue section using `plotVisium()`. Which cells are more likely to be excluded?
:::

:::{.callout-tip collapse="true"}
### Answer

```{r}
plotSpatialFeature(sfe, colGeometryName = "cellseg", features = "subsets_Mito_percent") +
  plotSpatialFeature(sfe, colGeometryName = "cellseg", features = "sum") +

plot_layout(nrow=1) &
    theme(plot.title=element_text(hjust=0.5)) &
    scale_fill_gradientn(NULL, colors=pals::jet())
```
:::

### Visualize spatial pattern of QC metrics (Global outliers)

Very importantly, we want to have a look into the spatial pattern of using a fixed threshold for QC metrics :

```{r}
# check spatial pattern of discarded cells
p1 <- plotObsQC(sfe, 
                plot_type="spot", annotate="qc_lib_size") + 
  ggtitle("Library size (< 400 UMI)")

p2 <- plotObsQC(sfe, 
                plot_type="spot", annotate="qc_detected") + 
  ggtitle("Detected genes (< 300 genes)")

p3 <- plotObsQC(sfe, 
                plot_type="spot", annotate="qc_mito_prop") + 
  ggtitle("Mito proportion (> 20%)")

# plot the three together
gridExtra::grid.arrange(p1, p2, p3, ncol=2)
```


### Local Outlier Detection

To avoid discarding whole layers of tissue, another approach to identify low-quality cells is to use local outlier detection methods that take into account the spatial context of each cell.

`SpotSweeper` package provides functions to identify common QC metrics (such as the ones we have seen in the previous section: library size, number of detected genes, and mitochondrial percentage) and detect local outliers based on these metrics.

We will use `localOutliers` from `SpotSweeper`, which helps identify cells that deviate significantly from their local neighborhood.


```{r}
# Identify local outliers based on library size ("sum" of counts).
# cells with unusually low library size compared to their neighbors will be flagged.
sfe <- localOutliers(sfe,
  metric = "sum",
  direction = "lower",
  log = TRUE
)

# Identify local outliers based on the number of unique genes detected ("detected").
# cells with an unusually low number of detected genes will be flagged.
sfe <- localOutliers(sfe, 
  metric = "detected",
  direction = "lower",
  log = TRUE
)

# Identify local outliers based on the mitochondrial gene percentage.
# cells with an unusually high mitochondrial percentage will be flagged.
sfe <- localOutliers(sfe,
  metric = "subsets_Mito_percent",
  direction = "higher",
  log = FALSE
)

# Combine all individual outlier flags into a single "local_outliers" column.
# A cell is considered a local outlier if it is flagged by any of the above metrics.
sfe$local_outliers <- as.logical(sfe$sum_outliers) |
  as.logical(sfe$detected_outliers) |
  as.logical(sfe$subsets_Mito_percent_outliers)
```

::: callout-important
## Exercise 3
How many cells were identified as local outliers based on the combined criteria?

As we have seen before, visualizing the QC metrics is crucial for understanding the quality of your spatial transcriptomics data and for deciding on appropriate filtering strategies, so look at their spatial localization using the `plotObsQC` function.

What do you think?
:::


:::{.callout-tip collapse="true"}
### Answer
```{r}
# Count the number of cells flagged as local outliers.
table(sfe$local_outliers)

plotObsQC(sfe, 
          plot_type="spot", annotate="local_outliers") + 
  ggtitle("All local outliers")
```


Even better, we can visualize the spatial distribution of the local outliers together with the QC metrics, for example plot the spatial distribution of local outliers for mitochondrial percentage.
```{r}
plotQCmetrics(sfe,
  metric = "subsets_Mito_percent", outliers = "subsets_Mito_percent_outliers",
  point_size = 0.4,
  stroke = 0.75
) +
  ggtitle("Local outliers for %MT")
```
:::

::: callout-important
## Exercise 4
Plot the spatial distribution of local outliers for library size and number of detected genes. What do you observe?
:::

:::{.callout-tip collapse="true"}
### Answer

```{r}
# Plot the spatial distribution of local outliers for library size.
plotQCmetrics(sfe,
  metric = "sum_log", outliers = "sum_outliers", point_size = 0.4,
  stroke = 0.75
) +
  ggtitle("Local outliers for number of UMIs")
```

```{r}
# Plot the spatial distribution of local outliers for number of detected genes.
plotQCmetrics(sfe,
  metric = "detected_log", outliers = "detected_outliers", point_size = 0.4,
  stroke = 0.75
) +
  ggtitle("Local outliers for number of UMIs")
```

Very few cells seem to be flagged as outliers!
:::


We can visualize the z-scores from the local outlier detection test and compare them to the global threshold used earlier. How do you interpret this?
```{r}
plotObsQC(sfe, plot_type = "violin", x_metric = "subsets_Mito_percent_z", annotate = "qc_mito_prop")
```

## Compare and combine global and local outliers and apply filtering

Based on the comparison of global and local outliers, we choose here to adjust the global cutoffs to be less stringent:

Adjust global filtering cutoffs based on previous plots (apply less stringent cutoffs)

```{r}
# adjust global filtering cutoff
sfe$qc_lib_size <- sfe$sum < 120
sfe$qc_detected <- sfe$detected < 100
sfe$qc_mito_prop <- sfe$subsets_Mito_percent > 35
```

Compare global and local outliers being flagged

```{r}
sfe$global_outliers <- sfe$qc_lib_size | sfe$qc_detected | sfe$qc_mito_prop
p1 <- plotObsQC(sfe, plot_type="spot", annotate="global_outliers")
p2 <- plotObsQC(sfe, plot_type="spot", annotate="local_outliers")
gridExtra::grid.arrange(p1, p2, ncol=2)
```

Combine both:

```{r}
# combine local and global outliers to create a final discard flag
sfe$discard <- sfe$global_outliers | sfe$local_outliers
```

::: callout-important
## Exercise 5
Visualize the cells that will be discarded based on the combined criteria. How many cells will be removed?
:::

:::{.callout-tip collapse="true"}
### Answer
```{r}
# check how many cells will be discarded
table(sfe$discard)

# have a final look at the cells that we will discard
plotSpatialFeature(sfe, colGeometryName = "cellseg", features = "discard")
```
:::

## Technical artifacts

<!--- JUNA: maybe we could skip the technical artifacts part, the result is not easy to interpret in this section -->

Technical artifacts can arise during tissue preparation, for example tissue dissection can cause mechanical damage, leading to large areas with artificially low biological heterogeneity. These can be detected with the `findArtifacts` function of the `spotSweeper` package.

```{r}
# find artifacts using SpotSweeper
sfe <- tryCatch(
  findArtifacts(sfe,
    mito_percent = "subsets_Mito_percent",
    mito_sum = "subsets_Mito_sum",
    n_order = 5,
    name = "artifact"
  ),
  error = function(e) {
    message(
      "findArtifacts() could not fit an artifact model for this subset: ",
      conditionMessage(e)
    )
    sfe$artifact <- FALSE
    sfe
  }
)
```

::: callout-important
## Exercise 6
What is the `findArtifacts` function performing and what are the (potentially problematic) assumptions?
:::

::: callout-important
## Exercise 7
Plot the spatial distribution of the `artifact` column identified by `SpotSweeper`. What do you observe?
:::

:::{.callout-tip collapse="true"}
### Answer

```{r}
# Plot the spatial distribution of the detected artifacts.
plotQCmetrics(sfe,
    metric = "subsets_Mito_percent", # You can choose any relevant metric to display
    outliers = "artifact", 
    point_size = 1.1
) +
    ggtitle("Detected Spatial Artifacts")
```
:::

## Filtering 
We decide to filter out the low-quality cells from the `SpatialExperiment` object based on the combined criteria of global thresholds and local outliers. Additionally, we will remove any features (genes) that have zero counts across all remaining cells.

```{r}
# remove the cells considered of low quality based on both global and local metrics
sfe <- sfe[, !sfe$discard]

# remove features with all 0 counts
sfe <- sfe[rowSums(counts(sfe)) > 0, ]
``` 

What is the size of the dataset after filtering?

## Save the object

Save the filtered `SpatialExperiment` object for the next steps:
```{r}
saveHDF5SummarizedExperiment(sfe, dir="results/day1", prefix="01.2_sfe_", replace=TRUE,
                                chunkdim=NULL, level=NULL, as.sparse=NA,
                                verbose=NA)
```

Clear your environment:
```{r}
#| eval: false
rm(list = ls())
gc()
.rs.restartR()
```

:::{.callout-important}
**Key Takeaways:**

-   `scuttle::addPerCellQCMetrics` provides essential per-bin QC metrics.
-   `SpotSweeper::localOutliers` helps identify individual problematic cells.
-   `SpotSweeper::findArtifacts` is powerful for detecting spatially coherent regions of poor quality.
-   Visualizing QC results is critical for data interpretation and downstream analysis decisions.
:::
