---
title: "Exercise 7"
format: html
editor: source
editor_options: 
  chunk_output_type: console
execute:
  warning: false
---

```{r}
#| message: false
#| warning: false
#| output: false

library(patchwork)
library(SpatialExperiment)
library(SpatialFeatureExperiment)
library(Voyager)
library(DropletUtils)
library(SingleR)
library(viridis)
library(RColorBrewer)
library(pheatmap)
library(alabaster.sfe)
```

## Dataset loading
We will start by loading the segmented Visium HD dataset stored in the `SpatialFeatureExperiment` object generated in the previous exercise. It contains information on the non-spatially-aware and spatially-aware clusters computed with graph-based clustering Banksy and PCA.

```{r}
# Reload the SpatialFeatureExperiment object if not in the R session already
sfe <- readObject("results/day2/02.3_sfe/")
sfe
```

## Perform cell-type annotation of the cells

To transfer cell-type labels to out Visium HD dataset, we'll use a matching scRNA-seq dataset generated for the same article by the 10X team, see the record on [GEO](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE280311).

::: callout-important
## Exercise 1
Which technology was used to generate this dataset anf what is the design of the experiment?
:::

::: {.callout-tip collapse="true"}
### Answer
This scRNA-seq was generated with the 10X FLEX technology, which is probe-based and allows to work with FFPE tissue blocks.

Sample barcodes, sequended together with the probe sequences allow to pool multiple samples together and then demultiplex them. Here 2 pools of 2 samples (P2CRC/P4CRC, P3NAT/P5NAT) and 1 pool of 4 samples (P1CRC/P3CRC/P5CRC/P2NAT) were created and each pool was sequenced on multiple 10X lanes
:::

### Load the FLEX dataset

The dataset should already have been downloaded with the Visium HD and Xenium data at the beginning of the course

```{r}
ref.sce <- DropletUtils::read10xCounts("data/Human_Colon_Cancer_P2/chromium/Human_DTC_5pv2_P5_CRC_count_filtered_feature_bc_matrix.h5")
## TO DO: this doesn't seem to be the same dataset... Plus we need the annotated data!




```

### Perform the anotation with `SingleR`
To annotate each cell, we use the `SingleR()` method, which uses a correlation-based approach comparing test cells (here: cells) to reference samples with known cell-type annotation.


```{r}
table(colData(ref.sce)$labels)

# Predict cell-type using SingleR 
pred <- SingleR(test   = logcounts(sfe),
                ref    = assay(ref.sce, "data"), # log-normalized counts        
                labels = ref.sce$labels,
                aggr.ref = TRUE,                # pseudo-bulk aggregation of reference labels  
                fine.tune = FALSE,              # for faster computation here
)
sfe$SingleR_label <- pred$pruned.labels
table(sfe$SingleR_label)
# visualize the predictions on the tissue slide.
plotCoords(sfe, annotate="SingleR_label", point_shape = 15, point_size = 1) + 
  scale_colour_brewer(palette = "Set1")
plotVisium(sfe, annotate="SingleR_label", zoom = T, point_shape = 22, point_size = 1) + 
  scale_fill_brewer(palette = "Set1")
```

:::{.callout-important}
## Exercise 2
What do you think about this annotation? Do you identify potentially relevant biological structures? How are these isolated above in the calculated clusters and spatial domains? 
:::

:::{.callout-tip collapse="true"}
### Answer

Let's compare to the Leiden cluster for example:
```{r}
tab_prop <- prop.table(table(sfe$Leiden, sfe$SingleR_label), margin = 1)
## For display: round and remove 0s
num_mat <- round(tab_prop, 2)
num_mat[num_mat == 0] <- ""

## Annotation colors
group_df = data.frame(row.names = colnames(tab_prop), SingleR_label=colnames(tab_prop))
ann_colors = list(
  SingleR_label = setNames(object = brewer.pal(9, "Set1"), colnames(tab_prop)))

pheatmap(
  tab_prop,
  cluster_rows = TRUE,
  cluster_cols = TRUE,
  color = viridis(100, option = "magma", direction = -1)[0:50],
  display_numbers = num_mat,
  fontsize_number = 10,
  fontsize = 12,
  border_color = NA,
  main = "Leiden clusters vs. SingleR labels", 
  annotation_col = group_df,
  annotation_colors = ann_colors  
)
```
:::
