---
title: "Exercise 7"
format: html
editor: source
editor_options: 
  chunk_output_type: console
execute:
  warning: false
---

```{r}
#| message: false
#| warning: false
#| output: false

library(SpatialExperiment)
library(SpatialFeatureExperiment)
library(Voyager)
library(DropletUtils)
library(SingleR)
library(tidyverse)
library(stringr)
library(patchwork)
library(viridis)
library(RColorBrewer)
library(pheatmap)
library(alabaster.sfe)
```

## Dataset loading
We will start by loading the segmented Visium HD dataset stored in the `SpatialFeatureExperiment` object generated in the previous exercise. It contains information on the non-spatially-aware and spatially-aware clusters computed with graph-based clustering Banksy and PCA.

```{r}
# Reload the SpatialFeatureExperiment object if not in the R session already
sfe <- readObject("results/day2/02.3_sfe/")
sfe
```

To transfer cell-type labels to out Visium HD dataset, we will use a [matching scRNA-seq dataset](https://www.10xgenomics.com/platforms/visium/product-family/dataset-human-crc) generated by the 10X team for the same article. See the record on [GEO](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE280311).

::: callout-important
### Exercise 1
Which technology was used to generate the scRNA-seq dataset and how was the experiment designed?
:::

::: {.callout-tip collapse="true"}
### Answer
This scRNA-seq was generated with the 10X FLEX technology, which is probe-based and allows to work with FFPE tissue blocks.

Sample barcodes, sequenced together with the probe sequences allow to pool multiple samples together and then demultiplex them. Here 2 pools of 2 samples (P2CRC/P4CRC, P3NAT/P5NAT) and 1 pool of 4 samples (P1CRC/P3CRC/P5CRC/P2NAT) were created, so 8 samples in total, and each pool was sequenced on multiple 10X Chromium lanes.
:::

## Download the FLEX dataset

We will download the scRNA-seq data from GEO, but this doesn't include the cell metadata, in particular the annotated cell type that we need here. This is available from Github: https://github.com/10XGenomics/HumanColonCancer_VisiumHD/tree/main

```{r download}
options(timeout = 1000)

if (!dir.exists("data/FLEX/")) {
  download.file(
    url = "https://ftp.ncbi.nlm.nih.gov/geo/series/GSE280nnn/GSE280311/suppl/GSE280311_RAW.tar",
    destfile = "data/GSE280311_RAW.tar",
    mode = "wb"
  )
  untar(
    tarfile = "data/GSE280311_RAW.tar",
    exdir = "data/FLEX/"
  )
  file.remove("data/GSE280311_RAW.tar")

  ## Download the cell metadata from 10X Github page
  download.file(
    url = "https://raw.githubusercontent.com/10XGenomics/HumanColonCancer_VisiumHD/refs/heads/main/MetaData/SingleCell_MetaData.csv.gz",
    destfile = "data/FLEX/SingleCell_MetaData.csv.gz",
    mode = "wb"
  )
} else {
  message("Data exists, please proceed to next steps!")
}
```


## Load the P2CRC sample

The sample matching our Visium HD and Xenium dataset was sequenced on 4 10X channels. For simplicity here, we only load the cell from one channel.

```{r load_flex}
list.files("data/FLEX/", pattern = "P2CRC") 

ref.sce <- DropletUtils::read10xCounts("data/FLEX/GSM8594533_P1L1_P2CRC_BC1_count_filtered_feature_bc_matrix.h5")
ref.sce
colnames(ref.sce) <- ref.sce$Barcode

rownames(ref.sce) <- uniquifyFeatureNames(
  ID = rowData(ref.sce)$ID,
  names = rowData(ref.sce)$Symbol
)

## We will need the logcounts for the annotation
ref.sce <- logNormCounts(ref.sce)

ref.sce
```

::: callout-important
### Exercise 2
How many cells were sequenced in this specific run? Why are counts available for only 18,082 genes? What type of information is available in the `colData`?
:::

::: {.callout-tip collapse="true"}
### Answer
We loaded here 10,397 cells from one 10X channel run of the P2CRC sample (multiplexed with another channel).

The FLEX technology is probe-based so it doesn't cover the full transcriptome, mostly the protein-coding genes are targeted. 

So far, there is no information on the cells in the object, expect their 10X barcode.
:::

## Load the annotation

```{r load_annot}
cell_metadata <- read.csv("data/FLEX/SingleCell_MetaData.csv.gz")
dim(cell_metadata)
head(cell_metadata)

cell_metadata <- cell_metadata |>
  as_tibble() |>
  mutate(Sample = str_extract(Barcode, pattern = "(\\d)$")) |>
  filter(Patient == "P2CRC" & QCFilter == "Keep" & Sample == 1) 
table(colnames(ref.sce) %in% cell_metadata$Barcode) 

ref.sce <- ref.sce[, cell_metadata$Barcode]
colData(ref.sce) <- cbind(colData(ref.sce), cell_metadata)

## Explore the annotated object
data("ditto_colors")
ggcells(ref.sce, aes(x = UMAP1, y=UMAP2, color=Level1)) + 
  geom_point() +
  scale_colour_manual(values = ditto_colors) + 
  theme_bw() 
ggcells(ref.sce, aes(x = UMAP1, y=UMAP2, color=Level2)) + 
  geom_point() +
  scale_colour_manual(values = ditto_colors) + 
  theme_bw() 
```

::: callout-important
### Exercise 3
What do you observe?
:::

::: {.callout-tip collapse="true"}
### Answer
The cell types match the expectations based on the tissue sample we could study earlier in the course. It includes tumore cells, stromal cells and immune cells.

A few cells in the dataset could be misannotated, although this might be an impression induced by the distortion typical of dimensionality reductions like UMAP.
:::

## Perform the anotation with `SingleR`

To annotate each cell, we use the `SingleR()` method, which uses a correlation-based approach comparing cells from the Visium HD dataset to reference cells from the FLEX dataset with known cell-type annotation.

We will use below the highest-level annotation available. A pseudo-bulk aggregation step before the comparison allows for a better performance. We also disable the fine tuning setp for faster computation here:

```{r}
table(colData(ref.sce)$Level1)

# Predict cell-type using SingleR 
pred <- SingleR(test = sfe,
                ref = ref.sce,
                labels = ref.sce$Level1,
                aggr.ref = TRUE,       
                fine.tune = FALSE,     
)
sfe$SingleR_label <- pred$pruned.labels
table(sfe$SingleR_label)

# visualize the predictions on the tissue slide.
plotSpatialFeature(sfe,  
                   colGeometryName = "cellSeg",
                   features = "SingleR_label")

    
```

:::{.callout-important}
### Exercise 4
What do you think about this annotation? 

How do the proportions of different cell types compare to teh annotated scRNA-seq dataset?

Do you identify potentially relevant biological structures? Go back to the calculated clusters and spatial domains, to see how do they match 
:::

:::{.callout-tip collapse="true"}
### Answer
Let's calculate the proportion of cells from each cell type:
```{r}
prop.table(table(ref.sce$Level1))
prop.table(table(sfe$SingleR_label))
```
We observe a relatively higher proportion of Endothelial cells and Fibroblast in the Visium HD. Maybe such cell-types are less easily captured by the droplet-based scRNA-seq technology?

Let's compare the annotation to the Leiden clustering for example:
```{r}
plotSpatialFeature(sfe,  
                   colGeometryName = "cellSeg",
                   features = "SingleR_label") +
  plotSpatialFeature(sfe,  
                   colGeometryName = "cellSeg",
                   features = "Leiden")

## For a more formal comparison
tab_prop <- prop.table(table(sfe$Leiden, sfe$SingleR_label), margin = 1)
## Display the proportions: round and remove 0s
num_mat <- round(tab_prop, 2)
num_mat[num_mat == 0] <- ""

pheatmap(
  tab_prop,
  cluster_rows = TRUE,
  cluster_cols = TRUE,
  color = viridis(100, option = "magma", direction = -1),
  display_numbers = num_mat,
  main = "Leiden clusters vs. SingleR labels", 
  annotation_colors = list(SingleR_label = setNames(object = ditto_colors[1:9], colnames(tab_prop)))  
)
```
:::

:::{.callout-important}

### Exercise 5: Bonus

Perform the same annotation task on the Xenium dataset. How do the results compare? What could be a limitation here?
:::

:::{.callout-tip collapse="true"}
### Answer

One limitation of this Xenium dataset is the limited number of genes targeted by the probe panel.
:::
